Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2020-10-13 analysis/paper/Family_Fortunes_Article.Rmd
Final edits

abstract: 
   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*, or city-state. However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken. This paper provides such an analysis, examining both the synchronic and diachronic variation among the tombs in these two cemeteries. Through a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and considers the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups. This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age. It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use."

@@

# 1. Introduction
Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access, reproducible research. To begin with the first, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention. The city and palace of Knossos, first excavated well over a century ago [@Evans1921; -@Evans1928; -@Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@1950].

However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project (KULP) suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8]. This estimate remains our best guess in the absence of substantial excavated deposits but, if accurate, suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.

@@

Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods. Though many were robbed in antiquity, and others destroyed by the digging of later burials, building work, or just the ravages of time, they still represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.

@@

The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 (Brock 1957: xi), grouped into three main clusters running north to south. Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill [@Brock1957: 1-2], are now considered to belong the Knossos North Cemetery (KNC).

@@

During the use of the cemeteries, correspondences have been highlighted between changing pottery styles and episodes of tomb construction and abandonment. The Protogeometric B (PGB) style, first defined by Brock [-@Brock1957: 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b: 93-94]. Coldstream has argued that the concurrence of this style with a rash of newly founded tombs points to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T. 107; see @Coldstream1984b; -@Coldstream1994: 112–13]. Whether simply nostalgia [@Coldstream1998: 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994: 114-15], the Bronze Age past clearly possessed a significant and enduring cultural cachet for the EIA Knossians.

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA.

For instance, Tomb 201 in the KNC contained a bronze sword, spearhead, and arrowhead, an iron dirk or knife, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123]. Catling identifies similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi; all contained imported items, were of ‘warrior grave’ type, and formed loci for later burials. He has argued that individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, would have been accorded exceptional treatment, including the deposition of such items, in death [@Catling1995: 127–8].

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Variables include the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'oriental' imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016] respectively. In total, this database comprises just over 4540 individual artefacts.

@@

This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete. Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital (or even digitisable) data from a number of recent articles and monographs is less pardonable.

There are, of course, valid reservations regarding issues of copyright, the protection of sites from looting, the possibility of being scooped with one's own data, or simply the significant time investment of preparing open-access research. But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit hugely from the greater accessability of datasets and workflows.

In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [@Marwick2017; @inbook], an open-source package for the statistical software R Studio. This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the following analysis. Though not visible in the present paper, every chart and table herein was produced directly, through written code, from the raw data using R (and note also that, wherever possible, these are presented in colour-blind friendly palettes). This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and original article are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**). The data are made available through an MIT license, which permits unlimited use, ammendation, and dissemination thereof, yet frees the present author from any responsibilty for that use, or perceived errors or inaccuracies in the data. It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.

@@

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen broadly follows that of Cavanagh [-@Cavanagh1996: 659-60], who took the number of pithoi (the most common urn) and the number of pithoi, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit (Figure \@ref(fig:burialestimate)). Here, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.

@@

But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi); the apparent cessation of depositions in the 8th century is simply a fiction generated by the processing of the data.

@@

As seen in Figure \@ref(fig:aoristallcems), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations [@Crema2010: 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating.

@@

aorist_pithoi_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=11))+
            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

aorist_pots_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=11))+
            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+  
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations. In both cases, tombs lacking any securely dateable pottery are excluded. Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising"}

@@

date_by_vessel<-ggplot(data=subset(pots,pots$century!=""),aes(century,fill="Date"))+geom_bar()+
  theme_bw()+theme(legend.position="none",
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
                 labs(x="Century BC (Date of Vessel)", y="Vessel Count")+
                  scale_fill_manual(values = c("gray50"))+ 
                    scale_y_continuous(expand=c(0,0),limits=c(0,2800))

date_by_tomb<-ggplot(data=subset(pots,pots$century!=""),aes(tomb_century,fill="Date"))+geom_bar()+
  theme_bw()+theme(legend.position="none",
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
                 labs(x="Century BC (Date of Tomb)", y="Vessel Count")+
                  scale_fill_manual(values = c("gray50"))+ 
                    scale_y_continuous(expand=c(0,0),limits=c(0,2800))

@@

*Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries respectively. For each of the centuries the tombs were in use (column two), counts of pithoi (column three) are divided by the number of tombs from which these vessels came (column 4), allowing calculation of the changing rate of burial through time (column five).*

The significance of these findings can be seen in Table 1. Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time. And while these mostly remained around three per tomb per century, in the 8th and 7th centuries, tombs founded in the 9th century saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century. Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.

@@

The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)). In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries. All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.

@@

Interestingly, Tombs X, G, Q, 219 and 283 ('Group B') evince a different pattern, with their respective peaks coming around a century earlier, another period of increased burial activity. So despite receiving the greatest quantities of burials and grave goods, these tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab). It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community.

@@

```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of* dromos *length and chamber area. Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue."}

@@

Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation. All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*. But, later destruction notwithstanding, there appears nothing systematic about the variation visible here. If the size of one's tomb was intended to evoke or assert ideas of wealth or status, this appears to have occurred on a largely independent basis.

```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and* dromos *length) to the estimated total number of burials by tomb. The points area colour-coded as per Figure 11."}

tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
  theme_bw()+theme(text=element_text(family="serif", size=10))+
            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
  theme_bw()+theme(text=element_text(family="serif", size=10))+
            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional. They clearly kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two hundred years. Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined. That is, these tombs seem to have followed a more conventional trajectory viz. Figure \@ref(fig:tombsinusebybuilddate). What might account for this disparity? And if the patterns implied in Figure \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in depositions - are genuine, how might we further investigate their manifestations across the excavated assemblage?

@@

```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Boxplot showing the aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period. Boxes represent the interquartile range, horizontal lines the median value, whiskers values within 1.5 times the interquartile range, and points outliers beyond this. Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively."}

@@

Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums across both cemeteries, colour-coded by ceramic phase. Looking at the PGB and EG, we see that they have the largest interquartile ranges (IQRs) of any period, and only a single outlier (beyond 1.5 times the IQR) each. The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (namely, Group A) far and above the range of the other tombs. In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume of burial activity, in the earlier period this was distributed far more evenly across all tombs. One speaks of increasing plurality, the other of divergence.

@@

This is not to suggest that the late 9th to early 8th century#ies was a time of egalitarian burial practices. The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12]. The PGB style reflects a uniquely Cretan blend of Attic and Near Eastern forms, while the appearance of BA 'antiques' in the tombs from this period onwards is felt to reflect rejuvinated, even competitive interest in the Minoan past.

@@

#####pagebreak

### Colophon

>>>>

abstract: 
   "The Early Iron Age cemeteries of Knossos, Crete, represent the richest source of archaeological material at the site for the whole period between the end of the Bronze Age and the emergence of the Archaic *polis*, or city-state. However, despite well-published excavation reports on the two largest burial grounds and a number of studies dealing with various artefacts and individual tombs therein, no substantial comparative, quantitative analysis has yet been undertaken. Such an analysis is provided here, examining both the synchronic and diachronic variation among the tombs of these two cemeteries. Through a range of quantitative techniques, this study isolates a pattern of increased rates of burial among a few select tombs, and considers the broader social contexts within which these tombs were built and used, as well as discussing the possible nature of the burying groups. This paper demonstrates the potential for quantitative analyses to contribute to nuanced, context-specific theories of social practice, and advocates their wider application to the archaeological record of the Cretan Early Iron Age. It also serves as a demonstration of maximal research reproducibility, with the full datasets and code underpinning the analysis made freely available to download and use."

@@

# 1. Introduction
Three principal interests converge in this paper: a reorientation of the focus of Early Iron Age Cretan archaeology; the application of computational, and quantitative methods to supplement traditional modes of archaeological inference; and open-access, reproducible research. To begin with the first, it may be felt that, of all ancient sites on Crete, Knossos is the last in need of further archaeological attention. The city and palace of Knossos, first excavated well over a century ago [@Evans1921; @Evans1928; @Evans1930], retain a pre-eminent position in popular and academic accounts the island’s Bronze Age (BA), and few, if any, other Cretan sites have been so thoroughly dug, reconstructed, and contested.

@@

Settlement evidence at EIA Knossos is, in a word, insubstantial. Speaking of the town’s stratigraphic sequence, Coldstream lamented that its “remains [...] are like a meagre filling in a very thick sandwich” between Bronze Age and later Greek and Roman levels [@Coldstream1991: 287]. Most individual finds of Protogeometric (PG) or Geometric date take the form of flimsy foundations, patches of earth floor, wells or pottery scatters, and for a time it was argued that the area in the EIA comprised several dispersed villages, which later coalesced as per Aristotle’s model of synoecism [@Alexiou1950].

However, a concentration of activity slightly to the west of the palace presented a good case for continuing, nucleated settlement [@Hood1981], something strongly advocated by Coldstream [-@Coldstream1984a; -@Coldstream2000], and, more recently, findings of the Knossos Urban Landscape Project (KULP) suggest a PG settlement of up to 40ha [@Kotsonas2011a: 5-8]. If accurate, this suggests that EIA Knossos was one of the largest sites of the contemporary Aegean.

@@

Fortunately, far larger burial accumulations are to be found in the Fortetsa and the Knossos North cemeteries, which together comprise some 134 tombs, ranging in date from the Subminoan (SM) to Late Orienatalising (LO) periods. TAken together, they represent the most abundant source of evidence for not just mortuary, but social practice of any kind at EIA Knossos.

@@

The cemetery on the slope facing the eponymous village of Fortetsa comprises twelve tombs excavated in 1933, and another eight uncovered in 1935 [@Brock1957: xi, 1-2], grouped into three main clusters running north to south. Three of those excavated in 1933 – L, TFT and Π – which lie close to the main road north of the Acropolis hill , are now considered to belong the Knossos North Cemetery (KNC).

@@

@@

A popular approach to the study of Knossos’ EIA tombs has been the investigation of ‘object biography’ [@Appadurai1986; @Kopytoff1986; @Mauss1954], an anthropological lens emphasising the mutuable nature of an object’s value, the transformations latent within its social existence. Fertilised by the Homeric motif of gift-giving [@Whitley2002: 220-21], and the remarkable archaeological parallels for such items – such as the boar’s tusk helmet from KNC Tomb 201 and that given to Odysseus by Meriones (*Iliad* 10.260-271) - such approaches have been brought to bear on certain elaborate burials of the EIA. For instance, Tomb 201 in the KNC contained various weapons, fragments of an antique bronze stand, probably of Cypriot origin, and pieces of the aforementioned helmet [@Catling1995: 123]. Based similarities between this tomb and burials at Tiryns, Kaloriziki, and the famous *hērōön* of Lefkandi, Catling has argued that these burials represent individuals who dared to travel abroad in this time of more limited seafaring, accruing exotica and stories from distant lands, and accorded exceptional treatment in death [@Catling1995: 127–8].

@@

The cornerstone of the present study is a database (or, rather, three linked databases) based on the excavation reports for the Fortetsa and Knossos North cemeteries [@Brock1957; @Coldstream1996a]. The first includes an entry for each individual ceramic vessel. Variables include the tomb, the vessel shape, its surviving dimensions, its ceramic period and corresponding absolute dating measures. In addition, all imports, 'Oriental' (read, Cypriot or Near Eastern) imitations, and antique (that is, Bronze Age) artefacts were categorised, using the studies of Antoniadis (2012), Jones [-@Jones2000], and Crowe [-@Crowe2016] respectively. In total, this database comprises just over 4540 individual artefacts.

@@

This article is, among other things, an attempt to bring these principles of maximal reproducibility to the archaeology of EIA Crete. Within this field, catalogues and excavation reports abound and, though many of these may be forgiven based on their dates of publication, the lack of readily utilisable digital (or even digitisable) data from a number of recent articles and monographs is less pardonable. There are, of course, valid reservations regarding the risks involved in sharing  archaeological data, not to mention the significant time investment of preparing open-access research. But I would argue that the potential gains outweigh these concerns, and that research on the EIA would benefit hugely from the greater accessability of datasets and workflows.

In this vein, the present article has been composed making use of Ben Marwick's 'rrtools' [@Marwick2017; @inbook], an open-source package for the statistical software R Studio. This provides the author with the tools to create a self-contained 'compendium', which can be freely downloaded, and from which anyone can reproduce the entire workflow of the published analysis. Though not visible in the present paper, every graph and table herein was produced directly, through written code, from the raw data using R (and note also that, wherever possible, these are presented in colour-blind friendly palettes). This code, as well as the raw and derived datasets, complete bibliography (in BibTex format, for export to any major reference manager), figures, and original article are included in the compendium. It can be downloaded as a repository from GitHub, at the following address: https://github.com/DCPollard94/knossoscemeteries, and the raw datasets have been stored with the online repository Figshare, here: (**NB a URL will be included pending acceptance of the article**). It is sincerely hoped that others may further explore and expand upon the research presented here by downloading this compendium.

@@

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen broadly follows that of Cavanagh [-@Cavanagh1996: 659-60], who took the number of pithoi (the most common urn) and the number of pithoi, amphorae and kraters (which were sometimes used as such) as middle and upper estimates of the number of burials, with all securely identified interments being the lower limit. Here, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis.

@@

But the artificial clustering this technique engenders around select absolute dates can prove misleading, as in Figure \@ref(fig:allpithoi) (left); the apparent cessation of depositions in the 8th century is simply a fiction generated by the processing of the data.

@@

As seen in Figure \@ref(fig:allpithoi) (right), the artificial peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations [@Crema2010: 1123-24], they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating.

@@

aorist_pithoi_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_sum,fill=" All tombs  "))+
    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_sum,fill=" KNC  "))+
      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

aorist_pots_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_tombs,aes(decade,aoristic_sum,fill=" All tombs  "))+
    geom_col(data=aorist_KNC,aes(decade,aoristic_sum,fill=" KNC  "))+
      geom_col(data=aorist_FOR,aes(decade,aoristic_sum, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="serif", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
              scale_fill_manual(values = c("gray50","gray65","gray80"))+  
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

```{r tombconstruction, fig.dim=c(7.5,4), dpi=600, fig.cap="Tomb construction through time at both cemeteries. On the left, counts are based on the ceramic phase considered most likely to correspond to the tomb's first use (including simply 'PG', 'G' and 'O' where no greater specificity was possible). On the right, aoristic sums are presented, spreading the uncertainty associated with those general assignations. In both cases, tombs lacking any securely dateable pottery are excluded. Abbreviations: SM - Subminoan; (E/M/L)PG - (Early/Middle/Late) Protogeometric; PGB - Protogeometric B; (E/M/L)G - (Early/Middle/Late) Geometric; (E/L)O - (Early/Late) Orientalising."}

@@

date_by_vessel<-ggplot(data=subset(pots,pots$century!=""),aes(century,fill="Date"))+geom_bar()+
  theme_bw()+theme(legend.position="none",
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
                 labs(x="Date of Vessel (Century BC)", y="Vessel Count")+
                  scale_fill_manual(values = c("gray50"))+ 
                    scale_y_continuous(expand=c(0,0),limits=c(0,2800))

date_by_tomb<-ggplot(data=subset(pots,pots$century!=""),aes(tomb_century,fill="Date"))+geom_bar()+
  theme_bw()+theme(legend.position="none",
               text=element_text(family="serif", size=10),axis.text.y=element_text(angle=90, vjust=0, hjust=0.5))+
                 labs(x="Date of Tomb (Century BC)", y="Vessel Count")+
                  scale_fill_manual(values = c("gray50"))+ 
                    scale_y_continuous(expand=c(0,0),limits=c(0,2800))

@@

*Table 1 The number of pithoi (as a proxy for burials) deposited in tombs dating to the 10th, 9th, 8th and 7th centuries respectively. For each of the centuries the tombs were in use (column two), counts of pithoi (column three) are divided by the number of tombs from which these vessels came (column four), allowing calculation of the changing rate of burial through time (column five).*

The significance of these findings can be seen in Table 1. Using pithoi as a proxy for burials, we can estimate the changing rates of burial through time. And while these mostly remained around three per tomb per century, in the 8th and 7th centuries tombs founded in the 9th saw a rapid rise in the rate of burial, peaking at over 11 per tomb per century. Though no doubt rough estimates, the marked nature of this divergence suggests it is not simply an artefact of imperfect data.

@@

The 12 tombs with the most associated vessels are all chamber tombs, 10 dating to the 9th century and two to the 10th, and seven of them did exhibit late, intense episodes of deposition (Figure \@ref(fig:toptombs)). In fact, Tombs P, 292, 75, 107, 285, II and 218 (henceforth, 'Group A') between them account for `r round(calc1,digits=2)`% of the entire ceramic assemblage from both cemeteries. All but Tomb 285 date to the 9th century, and together they underpinned the discussed late spate of burials.

@@

Interestingly, Tombs X, G, Q, 219 and 283 ('Group B') evince a different pattern, with their respective peaks coming around a century earlier, another period of increased burial activity. So despite receiving the most burials and grave goods across both cemeteries, these 12 tombs nonetheless seem to have achieved that distinction via quite different histories, a divergence made clear in Figure \@ref(fig:groupsab). It is notable that tombs assinged to each group are seen in both cemeteries, suggesting from the outset that the differences between these may speak to social forces operating across the Knossian community.

@@

```{r tombgroupdimensions, fig.dim = c(3.5, 3.5),out.width="10%", dpi=600, fig.align='center', fig.cap="Scatterplot comparing the dimensions of *dromos* length and chamber area. Groups A and B, defined earlier, are illustrated and labelled in green and orange respectively, while all remaining tombs ('Group C') are in blue."}

@@

Looking more specifically at the tombs of Groups A and B, against the backdrop of the other 10th and 9th century tombs, there are perhaps hints of differentiation (Figure \@ref(fig:tombgroupdimensions)). All of them fall within the top 50% based on chamber area, and three have exceptionally long *dromoi*. But, later destruction notwithstanding, there appears nothing systematic about the variation visible here. If the size of one's tomb was intended to evoke or assert ideas of wealth or status, this appears to have occurred on a largely independent basis.

```{r tombdimensions, fig.dim = c(7.5, 4), dpi=600, fig.align='center', fig.cap="Scatterplots comparing tomb dimensions (chamber area and *dromos* length) to the estimated total number of burials by tomb. The points are colour-coded as per Figure 11."}

tombareas<-ggplot(subset(tombs,tombs$est_burials>1),aes(chamber_area,est_burials,color=group))+geom_point()+
  theme_bw()+theme(text=element_text(family="serif", size=11))+
            labs(x="Chamber Area (sq m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

dromoslengths<-ggplot(subset(tombs,tombs$est_burials>1),aes(dromos_length,est_burials,color=group))+geom_point()+
  theme_bw()+theme(text=element_text(family="serif", size=11))+
            labs(x="Dromos Length (m)", y="Estimated Total Burials")+
               scale_color_brewer(palette = "Set2",name="Group")+
                scale_x_continuous(expand=c(0,0),limits=c(0,10))+
                  scale_y_continuous(expand=c(0,0),limits=c(0,50))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

@@

And in Figure \@ref(fig:tombscompthroughtime), we see that the tombs of Group A, while not paultry in the 9th century, were by no means exceptional in the quantity of burial goods they received. They  kept pace with the other popular tombs of the time, but their major divergence only occurred in the subsequent two centuries. Group B, meanwhile, experienced similar, if not greater, use in the 9th century, and then gradually declined. That is, these tombs seem to have followed a more conventional trajectory viz. Figure \@ref(fig:tombsinusebybuilddate). What might account for this disparity? And if the patterns implied in Figure \@ref(fig:tombscompthroughtime) - relative diversity, giving way to effective bipolarity in depositions - are genuine, how might we further investigate their manifestations across the excavated assemblage?

@@

```{r bigbox, fig.dim=c(7,4),dpi=600, fig.align='center',fig.cap="Boxplot showing variation in aoristic sums of all vessels across both cemeteries through time, colour-coded by ceramic period. Note, the Y axis has been limited to 20 to aid readability, meaning the outliers Tomb G in the EG period, and Tomb P in the late EO-LO period are excluded, with values ranging from 29-34, and 36-44 respectively."}

@@

Figure \@ref(fig:bigbox) is a boxplot of the total aoristic sums per tomb across both cemeteries, colour-coded by ceramic phase. Looking at the PGB and EG, we see that they have the largest interquartile ranges (IQRs) of any period, and only a single outlier (beyond 1.5 times the IQR) each. The Orientalising period, on the other hand, exhibits much reduced IQRs, and a proliferation of outliers (namely, Group A) far and above the range of the other tombs. In other words, whilst both the PGB-EG and EO-LO periods witnessed an expanding volume of burial activity, in the earlier period this was distributed far more evenly across all tombs. One speaks of increasing plurality, the other of divergence.

@@

This is not to suggest that the late 9th to early 8th centuries was a time of egalitarian burial practices. The PGB and EG are consistently interpreted as periods of artistic development, increased overseas trade, and renewed vigour in the construction of tombs and the visibility of the burial rite [@Brock1957: 143-44; @Coldstream1994: 110-12]. The PGB style, first defined by Brock [-@Brock1957: 143], who hailed it as “the most remarkable phase in Cretan vase-painting”, is apparently a Cretan anomaly, evincing a melding of Attic geometric motifs, patterns possibly deriving from Near Eastern metalwork, and iconography drawn from the Minoan repertoire [@Coldstream1984b: 93-94]. Coldstream has tied the emergence of this style to a reformulation of elite mortuary display, noting how, in this period also, redeposited Minoan larnakes began appearing in and around some tombs, in one case even inspiring the decoration of a PGB vessel [in T. 107; see @Coldstream1984b; -@Coldstream1994: 112–13]. Whether simply nostalgia [@Coldstream1998: 60], or a more tactical strategy by newly emerging elite groups [@Coldstream1994: 114-15], the Bronze Age past appears to have possessed a significant cultural cachet for the EIA Knossians, as may the new, elaborately decorated pottery of the PGB, and items imported from overseas.

@@

##### pagebreak
### Colophon
